Methodology of Creating SVM Kernels from Scratch Using Python and NumPy

### 2023-12-15 Paper/paper.tex
Update

\begin{abstract}
This paper presents a methodology for creating Support Vector Machine (SVM) kernels from scratch using Python and NumPy. 
We discuss the implementation of linear, sigmoid, polynomial, and radial basis function (RBF) kernels in a binary and multiclass SVM.
These kernels are tested on E. coli data and compared to the results of the scikit-learn SVM implementation.
The goal of the E. coli dataset is to predict the localization sites of proteins inside the cell. 
The results show that the implemented kernels often yield better accuracy than the scikit-learn implementation; at the
cost of increased training time. Kernel training times are ran multiple times and averaged to provide a more accurate
metric since training times are low and have a high variance.
\end{abstract}

@@

Support Vector Machines (\textbf{SVMs}) are machine learning algorithms that are used to complete classification tasks. When we have a large dimensional data set that we would like to create decision boundaries, we can create a hyperplane that maximizes the margin between the observations' labels. The data points that are closest to the decision boundary influence the position and direction of the hyperplane, and these close data points are called \textbf{Support Vectors}. The goal is to optimize the distance between the support vectors and the hyperplane (also known as the \textbf{margin}), and so we can draw a line between the points.

With higher dimensional data, sometimes a linear boundary will not be effective, and so we can implement kernel tricks. These kernels allow for the original input to be mapped to a higher-dimensional space where a linear boundary can be drawn. We will demonstrate the usage of SVMs, as well as the effect that different kernels have on the same data.

@@

When in discussion of SVMs, we need to understand the steps behind the math to know what it is doing. Here, we will discuss how this algorithm is run to calculate our optimal parameters for our hyperplane decision boundary.

@@

We can combine this all to formulate our optimization problem: minimize $||w||$ constrained by $y_i (w \cdot x_i + b) \geq 1$ for $i = 1,...,N$.
We can set up a Lagrange dual optimization problem, such that:
$$L = \frac{1}{2}||w||^2 - \sum_{i=1}^{n} \alpha_i \cdot [y_i(w \cdot x + b) - 1]$$
Minimizing for $w$ and $b$, we get:
$$w = \sum_{i=1}^{n} \alpha_i y_i x_i$$
$$\sum_{i=1}^{n} \alpha_i y_i = 0$$

Our problem can now be in terms of a constrained minimization problem by plugging in the constraints we got. This can be simplified into a quadratic form. The final form of our problem will be

@@

A binary classification on the data is the foundation of the calculations. We want to form a hyperplane that separates two labels effectively, with a margin for error when making predictions on new observations. The steps to complete to find the terms for our hyperplane can be seen below:

@@

Using this algorithm, you are able to find the $w$'s and $b$ that define the hyperplane between the two classes. The $K()$ function represents the kernel that is being used, and so this is where the different transformations will have an effect.

\subsection{Multiclass SVM}
The multi-classification on a data set using SVM is only a slight variation from the binary classification. In binary, we trained the model on a \textbf{1 vs 1} basis. We had two classes, -1 and 1, and wanted to create a decision boundary between them. However, multiple classes means that we have more than two classes, and you can't just draw one line to separate the labels.

Therefore, the meta-algorithm for multi-class is to approach with a \textbf{one vs others} mindset. For however many classes you have, you want to create a decision boundary between one class and all the others, and repeating this until you have done so for each label. It follows the same steps as binary classification, but does it $s$ times for the number of labels in the data set to differentiate. 

@@

        Binary & sklearn & linear & 1.00000 & 0.00052 \\
        Binary & custom & linear & 0.99318 & 0.00795 \\
        Binary & sklearn & sigmoid & 0.68180 & 0.00066 \\
        Binary & custom & sigmoid & 0.99546 & 0.00757 \\
        Binary & sklearn & rbf & 1.00000 & 0.00054 \\
        Binary & custom & rbf & 0.99546 & 0.00977 \\
        Binary & sklearn & poly & 1.00000 & 0.00040 \\
        Binary & custom & poly & 0.97498 & 0.00929 \\
        Multi & sklearn & linear & 0.77940 & 0.00096 \\
        Multi & custom & linear & 0.81468 & 0.15420 \\
        Multi & sklearn & sigmoid & 0.47060 & 0.00183 \\
        Multi & custom & sigmoid & 0.82497 & 0.15641 \\
        Multi & sklearn & rbf & 0.75000 & 0.00156 \\
        Multi & custom & rbf & 0.82350 & 0.19177 \\
        Multi & sklearn & poly & 0.76470 & 0.00111 \\
        Multi & custom & poly & 0.83967 & 0.36326 \\

@@

For the binary SVMs, we selected the classes \textit{cp} and \textit{im} to be the positive and negative classes respectively.
We can see that the linear kernel has a linear decision boundary which is expected and the rest of the kernels have non-linear decision boundaries.
\begin{figure}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/linear_binary.png}
        \caption{Linear Binary SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/poly_binary.png}
        \caption{Polynomial Binary SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/rbf_binary.png}
        \caption{RBF Binary SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/sigmoid_binary.png}
        \caption{sigmoid Binary SVM}
    \end{subfigure}
    \caption{Binary SVM Decision Boundaries}
\end{figure}
\begin{figure}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/linear_multi.png}
        \caption{Linear Multi-Class SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/poly_multi.png}
        \caption{Polynomial Multi-Class SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/rbf_multi.png}
        \caption{RBF Multi-Class SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/sigmoid_multi.png}
        \caption{sigmoid Multi-Class SVM}
    \end{subfigure}
    \caption{Multi-Class SVM Decision Boundaries}
\end{figure}
\FloatBarrier 
\section{Conclusion}
This paper presented a methodology for creating Support Vector Machine (SVM) kernels from scratch using Python and NumPy.
We discussed the implementation of linear, sigmoid, polynomial, and radial basis function (RBF) kernels in a binary and multiclass SVM
and how Binary SVMs are used underneath the hood of multiclass SVMs. These kernels were tested on E. coli data and compared to the results
of the scikit-learn SVM implementation. The results show that the implemented kernels often yield better accuracy than the scikit-learn
implementation; at the cost of increased training time. Kernel training times were ran multiple times and averaged to provide a more
accurate metric as training times were low and had a high variance. By harnessing the power of dimensionality reduction, we were
able to show the decision boundaries in 2 dimensions while still retaining the accuracy of higher dimensional data. These 
decision boundaries show how each kernel can affect the shape of the classification boundary and how the kernels can be used to
classify non-linear data.

>>>>

\begin{abstract}
This paper presents a methodology for creating Support Vector Machine (SVM) kernels from scratch using Python and NumPy. 
We discuss the implementation of linear, sigmoid, polynomial, and radial basis function (RBF) kernels in a binary and multiclass SVM.
These kernels are tested on E. coli data and compared to the results of the scikit-learn SVM implementation.
The goal of the E. coli dataset is to predict the localization sites of proteins inside the cell. 
The results show that the implemented kernels will sometimes yield better accuracy than the scikit-learn implementation; at the
cost of increased training time. Kernel training times are run multiple times and averaged to provide a more accurate
metric since training times are low and have a high variance.
\end{abstract}

@@

Support Vector Machines (\textbf{SVMs}) are machine learning algorithms that are used to complete classification tasks. When we have a large dimensional data set that we would like to create decision boundaries, we can create a hyperplane that maximizes the margin between the observations' labels. The data points that are closest to the decision boundary influence the position and direction of the hyperplane, and these close data points are called \textbf{Support Vectors}. The goal is to optimize the distance between the support vectors and the hyperplane (also known as the \textbf{margin}), so we can draw a line between the points.

With higher dimensional data, sometimes a linear boundary will not be effective, and so we can implement kernel tricks. These kernels allow for the original input to be mapped to a higher-dimensional space where a linear boundary can be drawn. We will demonstrate the usage of SVMs and the effect that different kernels have on the same data.

@@

When discussing SVMs, we need to understand the steps behind the math to know what it is doing. Here, we will discuss how this algorithm calculates our optimal parameters for our hyperplane decision boundary.

@@

We can combine this all to formulate our optimization problem: minimize $||w||$ constrained by $y_i (w \cdot x_i + b) \geq 1$ for $i = 1,..., N$.
We can set up a Lagrange dual optimization problem, such that:
$$L = \frac{1}{2}||w||^2 - \sum_{i=1}^{n} \alpha_i \cdot [y_i(w \cdot x + b) - 1]$$
Minimizing for $w$ and $b$, we get:
$$w = \sum_{i=1}^{n} \alpha_i y_i x_i$$
$$\sum_{i=1}^{n} \alpha_i y_i = 0$$

Our problem can now be in terms of a constrained minimization problem by plugging in the constraints we have. This can be simplified into a quadratic form. The final form of our problem will be

@@

A binary classification of the data is the foundation of the calculations. We want to form a hyperplane that separates two labels effectively, with a margin for error when making predictions on new observations. The steps to complete to find the terms for our hyperplane can be seen below:

@@

Using this algorithm, you can find the $w$'s and $b$ that define the hyperplane between the two classes. The $K()$ function represents the kernel that is being used, and so this is where the different transformations will have an effect.

\subsection{Multiclass SVM}
The multi-classification on a data set using SVM is only a slight variation from the binary classification. In binary, we trained the model on a \textbf{1 vs 1} basis. We had two classes, -1 and 1, and wanted to create a decision boundary between them. However, multiple classes mean that we have more than two classes, and you can't just draw one line to separate the labels.

Therefore, the meta-algorithm for multi-class is to approach with a \textbf{one vs others} mindset. For however many classes you have, you want to create a decision boundary between one class and all the others and repeat this until you have done so for each label. It follows the same steps as binary classification, but does it $s$ times for the number of labels in the data set to differentiate. 

@@

        Binary & sklearn & linear & 1 & 0.00119 \\
        Binary & custom & linear & 0.99318 & 0.00719 \\
        Binary & sklearn & sigmoid & 0.6818 & 0.00069 \\
        Binary & custom & sigmoid & 0.95908 & 0.05421 \\
        Binary & sklearn & rbf & 1 & 0.00034 \\
        Binary & custom & rbf & 0.99546 & 0.00773 \\
        Binary & sklearn & poly & 1 & 0.00035 \\
        Binary & custom & poly & 0.95909 & 0.00926 \\
        Multi & sklearn & linear & 0.7794 & 0.00094 \\
        Multi & custom & linear & 0.81762 & 0.13374 \\
        Multi & sklearn & sigmoid & 0.4706 & 0.00199 \\
        Multi & custom & sigmoid & 0.68826 & 1.50449 \\
        Multi & sklearn & rbf & 0.8824 & 0.00111 \\
        Multi & custom & rbf & 0.82939 & 0.17472 \\
        Multi & sklearn & poly & 0.8529 & 0.00107 \\
        Multi & custom & poly & 0.84849 & 0.32641 \\

@@

For the binary SVMs, we selected the classes \textit{cp} and \textit{im} to be the positive and negative classes respectively.
We can see that the linear kernel has a linear decision boundary which is expected and the rest of the kernels have non-linear decision boundaries.
\begin{figure}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/linear_binary.png}
        \caption{Linear Binary SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/poly_binary.png}
        \caption{Polynomial Binary SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/rbf_binary.png}
        \caption{RBF Binary SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/sigmoid_binary.png}
        \caption{sigmoid Binary SVM}
    \end{subfigure}
    \caption{Binary SVM Decision Boundaries}
\end{figure}
\begin{figure}
    \centering
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/linear_multi.png}
        \caption{Linear Multi-Class SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/poly_multi.png}
        \caption{Polynomial Multi-Class SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/rbf_multi.png}
        \caption{RBF Multi-Class SVM}
    \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{plots/sigmoid_multi.png}
        \caption{sigmoid Multi-Class SVM}
    \end{subfigure}
    \caption{Multi-Class SVM Decision Boundaries}
\end{figure}
\FloatBarrier 
\section{Conclusion}
This paper presented a methodology for creating Support Vector Machine (SVM) kernels from scratch using Python and NumPy.
We discussed the implementation of linear, sigmoid, polynomial, and radial basis function (RBF) kernels in a binary and multiclass SVM
and how Binary SVMs are used underneath the hood of multiclass SVMs. These kernels were tested on E. coli data and compared to the results
of the scikit-learn SVM implementation. The results show that the implemented kernels can often yield better accuracy than the scikit-learn
implementation; at the cost of increased training time. Kernel training times were ran multiple times and averaged to provide a more
accurate metric as training times were low and had a high variance. By harnessing the power of dimensionality reduction, we were
able to show the decision boundaries in 2 dimensions while still retaining the accuracy of higher dimensional data. These 
decision boundaries show how each kernel can affect the shape of the classification boundary and how the kernels can be used to
classify non-linear data.
